\texttt{Davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-10-26 paper/old.tex
made change requested by PCF

\begin{figure}[tp] \centering
\includegraphics[width=0.6\textwidth]{figs/shareable_code_2d} \caption{\small
\textbf{Systems for ensuring code reproducibility within the Python ecosystem.}
The $x$-axis denotes the ``burden'' placed on users to install and configure
the given system (systems placed further to the right, that fall within the
redder shading, impose a higher setup cost on the user). The $y$-axis denotes
the degree to which the system guarantees that the code will run similarly for
different users (systems placed higher up, that fall within the bluer shading,
offer stronger guarantees). From left to right, bottom to top: plain-text
\textbf{Python scripts} (\texttt{.py} files) provide the most basic ``system''
for sharing raw code. Scripts may reference external packages, but those
packages must be manually installed on other users' systems. Further, any
checking needed to verify that the correct versions of those packages were
installed must also be performed manually. \textbf{Jupyter notebooks}
(\texttt{.ipynb} files) comprise embedded text, executable code, and media
(including rendered figures, code output, etc.). When the \textbf{Davos
package} is imported into a Jupyter notebook, the notebook's functionality is
extended to automatically install any required external packages (at their
correct versions, when specified) into an isolated directory kept separate from
the user's existing set of packages. \textbf{Virtual environments} (along with
dependency specification files) similarly allow users to install required
packages into an isolated directory, and \textbf{Python environments} (again,
along with dependency specification files) extend this to include a specific
version of Python itself. However, both of these systems require users to
create, populate, and manage environments manually. This typically entails
distributing a configuration file (e.g., a \texttt{requirements.txt},
\texttt{pyproject.toml}~\cite{CannEtal16}, or \texttt{environment.yml} file)
that specifies all project dependencies (including version numbers) alongside
the primary code base. Users can then install a third-party
tool~\cite[e.g.,][]{BickEtal07, Eust19, Anac12} to read the file and build the
environment. \textbf{Containers} provide a means of defining an isolated
environment that includes a complete operating system (independent of the
user's operating system), in addition to (optionally) specifying a virtual
environment or other configurations needed to provide the necessary computing
environment. Containers are typically defined using specification files (e.g.,
a plain-text \texttt{Dockerfile}~\cite{Merk14}) that instruct the
virtualization engine regarding how to build the containerized environment.
\textbf{Virtual machines} provide a complete hardware-level simulation of the
computing environment. In addition to simulating specific hardware, virtual
machines (typically specified using binary image files) must also define
operating system-level properties of the computing environment. $\Delta_1$
represents the increased reproducibility \texttt{Davos} provides over standard
Jupyter notebooks for no greater setup cost. $\Delta_2$ represents
\texttt{Davos}'s lower setup cost compared to standard virtual environments
despite more stringently ensuring reproducibility.} \label{fig:code-sharing}
\end{figure}

@@

At another extreme, virtual machines~\cite{Gold74, AltiEtal05, Rose99} provide
a hardware-level simulation of the desired system. Virtual machines are
typically isolated, such that installing or running software on a virtual
machine does not impact the user's primary operating system or computing
environment. Containers~\cite[e.g.,][]{Merk14, KurtEtal17} provide a similar
``isolated'' experience. Although containerized environments do not specify
hardware-level operations, they are typically packaged with a complete
operating system, in addition to a complete copy of Python and any relevant
package dependencies. Shareable Python environments~\cite[e.g.,][]{Anac12} also
provide a computing environment that is largely separated from the user's main
environment. They incorporate a copy of Python and the target software's
dependencies, but do not specify or reproduce an operating system for the
runtime environment. Virtual environments~\cite[e.g.,][]{BickEtal07, Eust19}
work similarly, but reuse an existing copy of Python rather than bundling their
own. Each of these systems (virtual machines, containers, Python environments,
and virtual environments) guarantees (to differing degrees---at the hardware
level, operating system level, Python environment level, and package
environment level, respectively) that the relevant code will run similarly for
different users. However, each of these systems either relies on additional
software that can be complex or resource-intensive to install and use, or
requires additional setup, creating potential barriers to both contributing to
and taking advantage of open science resources. For example, what is typically
``shared'' with the end user in these cases is a set of instructions or
configuration files for \textit{building} the desired computing environment,
rather than the environment itself.

We designed \texttt{Davos} to occupy a ``sweet spot'' within this space.
\texttt{Davos} is a notebook-installable package that adds functionality to the
default notebook experience. Like standard Jupyter notebooks,
\texttt{Davos}-enhanced notebooks allow researchers to include text, executable
code, and media within a single file. No further setup or installation is
required from the user, beyond what is needed to run a standard Jupyter
notebook. And like separate dependency specification files used to build
virtual environments or Python environments, \texttt{Davos} provides a
convenient mechanism for fully specifying (and installing, as needed) a
complete set of Python dependencies, including specific package versions, which
are contained and isolated from the rest of the user's system.

@@

While the \texttt{Davos} package by no means offers a universal solution to this problem, it can, in many cases, provide an effective yet more accessible alternative to existing systems for sharing reproducible Python code, such as specification files for building virtual environments, Python environments, containers, and virtual machines (Fig.~\ref{fig:code-sharing}).
For researchers, this can lower barriers to collaborating with peers and contributing to publicly available open science resources.
And by eliminating most of the setup costs of reconstructing the original researchers' computing environment(s), \texttt{Davos} also lowers barriers to entry for members of the scientific community and the public who seek to run shared code.

Among common systems for sharing reproducible Python code (see Fig.~\ref{fig:code-sharing}), \texttt{Davos} is most comparable to specification files for building virtual environments in that it provides a lightweight mechanism for specifying and sharing a complete set of packages (and specific package versions) required by a particular project, and installing those packages into an isolated directory on the user's file system.
However, when used in conjunction with Jupyter notebooks, \texttt{Davos} offers a number of advantages over standard virtual environments that make it both easier to use and more effective at ensuring reproducibility of shared code.

>>>>

\begin{figure}[tp]
\centering
\includegraphics[width=0.65\textwidth]{figs/shareable_code_2d}
\caption{\small \textbf{Systems for sharing code within the Python
  ecosystem.} The $x$-axis denotes the ``burden'' placed on users to
  install and configure the given system (systems placed further to the right, that fall within the redder shading, impose a higher setup cost on the user). The $y$-axis denotes the
  degree to which the system guarantees that the code will run
  similarly for different users (systems placed higher up, that fall within the bluer shading, offer stronger guarantees).
  From left to right, bottom to top: plain-text \textbf{Python
  scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code. Scripts may reference external packages, but
  those packages must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those packages were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.). When the \textbf{Davos package} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install any required external packages (at their
  correct versions, when specified) into an isolated directory kept separate
  from the user's existing set of packages. \textbf{Virtual environments}
  similarly allow users to install required packages into an isolated
  directory, and \textbf{shareable Python environments} extend this to include
  a specific version of Python itself. However, both of these systems require
  users to create, populate, and manage environments manually. This
  typically entails distributing a configuration file (e.g., a
  \texttt{requirements.txt}, \texttt{pyproject.toml}~\cite{CannEtal16}, or
  \texttt{environment.yml} file) that specifies all project dependencies
  (including version numbers) alongside the primary code base. Users can then
  install a third-party tool~\cite[e.g.,][]{BickEtal07, Eust19, Anac12} to read
  the file and build the environment. \textbf{Containers} provide a means of
  defining an isolated environment that includes a complete operating
  system (independent of the user's operating system), in addition to
  (optionally) specifying a virtual environment or other
  configurations needed to provide the necessary computing
  environment. Containers are typically defined using
  specification files (e.g., a plain-text \texttt{Dockerfile}~\cite{Merk14}) that instruct the
  virtualization engine regarding how to build the containerized
  environment. \textbf{Virtual machines} provide a complete
  hardware-level simulation of the computing environment. In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary image files) must also define operating
  system-level properties of the computing environment. $\Delta_1$
  represents the increased reproducibility \texttt{Davos} provides over standard
  Jupyter notebooks for no greater setup cost. $\Delta_2$ represents
  \texttt{Davos}'s lower setup cost compared to standard virtual environments
  despite more stringently ensuring reproducibility.}
\label{fig:code-sharing}
\end{figure}

@@

At another extreme, virtual machines~\cite{Gold74, AltiEtal05, Rose99}
provide a hardware-level simulation of the desired system. Virtual
machines are typically isolated, such that installing or running
software on a virtual machine does not impact the user's primary
operating system or computing environment.
Containers~\cite[e.g.,][]{Merk14, KurtEtal17} provide a similar
``isolated'' experience. Although containerized environments do not
specify hardware-level operations, they are typically packaged with a
complete operating system, in addition to a complete copy of Python
and any relevant package dependencies. Shareable Python
environments~\cite[e.g.,][]{Anac12} also provide a computing
environment that is largely separated from the user's main
environment. They incorporate a copy of Python and the target
software's dependencies, but do not specify or
reproduce an operating system for the runtime environment.
Virtual environments~\cite[e.g.,][]{BickEtal07, Eust19} work similarly, but
reuse an existing copy of Python rather than bundling their own.
Each of these systems (virtual machines, containers, Python environments, and virtual environments)
guarantees (to differing degrees---at the hardware level, operating
system level, Python environment level, and package environment level, respectively) that the
relevant code will run similarly for different users. However, each of
these systems also relies on additional software that can be complex
or resource-intensive to install and use, creating potential barriers
to both contributing to and taking advantage of open science
resources.

We designed \texttt{Davos} to occupy a ``sweet spot'' within this space.
\texttt{Davos} is a notebook-installable package that adds functionality to the
default notebook experience. Like standard Jupyter notebooks,
\texttt{Davos}-enhanced notebooks allow researchers to include text, executable
code, and media within a single file. No further setup or installation is
required from the user, beyond what is needed to run a standard Jupyter
notebook. And like virtual environments, \texttt{Davos} provides a convenient
mechanism for fully specifying (and installing, as needed) a complete set of
Python dependencies, including specific package versions, which are contained
and isolated from the rest of the user's system.

@@

While the \texttt{Davos} package by no means offers a universal solution to this problem, it can, in many cases, provide an effective yet more accessible alternative to existing systems for sharing reproducible Python code, such as virtual environments, containers, and virtual machines (Fig.~\ref{fig:code-sharing}).
For researchers, this can lower barriers to collaborating with peers and contributing to publicly available open science resources.
And by eliminating most of the setup costs of reconstructing the original researchers' computing environment(s), \texttt{Davos} also lowers barriers to entry for members of the scientific community and the public who seek to run shared code.

Among common systems for sharing reproducible Python code (see Fig.~\ref{fig:code-sharing}), \texttt{Davos} is most comparable to virtual environments in that it provides a lightweight mechanism for specifying and sharing a complete set of packages (and specific package versions) required by a particular project, and installing those packages into an isolated directory on the user's file system.
However, when used in conjunction with Jupyter notebooks, \texttt{Davos} offers a number of advantages over standard virtual environments that make it both easier to use and more effective at ensuring reproducibility of shared code.
